Well-quasi-orderings on word languages

### 2025-10-16 src/appendix.tex
fix: knowledges + refs + storytelling
Fix proof ordering in appendix to follow logical dependencies
Co-authored-by: AliaumeL <24816735+AliaumeL@users.noreply.github.com>
final version

\begin{proofof}{bi-infinite:lem}
    Let us assume that $L$ is infinite. The case when it is finite 
    is similar, but will result in a finite word.
    \todo{aliaume: do we need wqo here? the proof
      should go through without it: the sequence 
      $u_i$ is already increasing for infix}

@@

@@

>>>>

\begin{proofof}{bi-infinite:lem}
    Let us assume that $L$ is infinite. The case when it is finite 
    is similar, but will result in a finite word.

@@

\AP Let us briefly recall that a \intro{rational transduction} is a relation $R
\subseteq \Sigma^* \times \Gamma^*$ such that there exists a finite state
automaton that reads pairs of letters $(a,b) \in (\Sigma \cup \set{\epsilon})
\times (\Gamma \cup \set{\epsilon})$ and recognizes $R$. A class of languages
$\mathcal{C}$ is \intro{closed under rational transductions} if for every $L
\in \mathcal{C}$ and every rational transduction $R$, the language $R(L)
\defined \set{v \in \Gamma^* \mid \exists u \in L, (u,v) \in R}$ also belongs
to $\mathcal{C}$.

@@

\section{Proofs for Section~\ref{conclusion:sec}}

\begin{lemma}
  \label{dwclosed-wqo-complexity-width-finite:lem}
  Let $L$ be a \kl{downwards closed} language that is
  \kl{well-quasi-ordered} by the \kl{infix relation}.
  Then, the following are equivalent:
  \begin{enumerate}
    \item $L$ has bounded \kl{factor complexity},
      \label{dwclosed-wqo-complexity-width-finite:item1}
    \item $L$ has finite \kl{ordinal width}.
      \label{dwclosed-wqo-complexity-width-finite:item2}
    \item $L$ is a finite union 
      of languages of the form
      $\infset{w}$ where $w$
      is an \kl{ultimately periodic} word.
      \label{dwclosed-wqo-complexity-width-finite:item3}
  \end{enumerate}
\end{lemma}
\begin{proof}
  \cref{dwclosed-wqo-complexity-width-finite:item3} $\Rightarrow$
  \cref{dwclosed-wqo-complexity-width-finite:item1} is clear because
  \kl{ultimately periodic} words have bounded \kl{factor complexity}.

  \cref{dwclosed-wqo-complexity-width-finite:item1} $\Rightarrow$
  \cref{dwclosed-wqo-complexity-width-finite:item2} is also clear because
  unbounded \kl{factor complexity} implies the existence of arbitrarily
  large \kl{antichains}.

  Finally, \cref{dwclosed-wqo-complexity-width-finite:item2} $\Rightarrow$
  \cref{dwclosed-wqo-complexity-width-finite:item3} is a direct consequence
  of \cref{dw-closed-wqo-charac:thm} and the fact that bounded \kl{factor
  complexity} implies that the (bi)infinite words describing the language
  are \kl{ultimately periodic}.
\end{proof}

### 2025-10-16 src/conclusion.tex
final version

>>>>

To conclude on a positive note for the \kl{infix relation}, our results show
that for \kl{downwards closed} and \kl{well-quasi-ordered} languages, there is
a strong connection between the \kl{factor complexity} and the \kl{ordinal
width}: it is the same to have bounded \kl{factor complexity} and finite
\kl{ordinal width}. A short proof can be found in appendix 
(\cref{dwclosed-wqo-complexity-width-finite:lem}).

### 2025-10-16 wqo-on-words.lncs.tex
final version

\usepackage{todonotes}

@@

% links inside the document
\usepackage{hyperref}
\usepackage[capitalise,noabbrev,nameinlink]{cleveref}
\Crefname{assumption}{Assumption}{Assumptions}
\Crefname{conjecture}{Conjecture}{Conjectures}
\usepackage[composition,hyperref,xcolor,cleveref]{knowledge}
\knowledgeconfigure{notion}

@@

\begin{document}
%
\title{Well-quasi-orderings on word languages \\
       main 81b14ea82bdf369ba5dbe8de29fa418756900125 \\
       2025-10-15 17:43:32 +0200
}
\titlerunning{WQO on Words}

@@

%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    The set of finite words over a well-quasi-ordered set is itself well-quasi-ordered. This seminal result by Higman is a cornerstone of the theory of well-quasi-orderings and has found numerous applications in computer science. However, this result is based on a specific choice of ordering on words, the (scattered) subword ordering. In this paper, we describe to what extent other natural orderings (prefix, suffix, and infix) on words can be used to derive Higman-like theorems. More specifically, we are interested in characterizing \emph{languages} of words that are well-quasi-ordered under these orderings, and explore their properties and connections with other language theoretic notions. We furthermore give decition procedures when the languages are given by various comupational models such as automata, context-free grammars, and automatic structures.
\end{abstract}

>>>>

@@

% links inside the document
\usepackage{hyperref}
\usepackage[capitalise,noabbrev,nameinlink]{cleveref}
\Crefname{assumption}{Assumption}{Assumptions}
\Crefname{conjecture}{Conjecture}{Conjectures}
\usepackage[electronic,hyperref,xcolor,cleveref]{knowledge}
\knowledgeconfigure{notion}

@@

\begin{document}
%
\title{Well-quasi-orderings on word languages}
\titlerunning{WQO on Words}

@@

%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    The set of finite words over a well-quasi-ordered set is itself well-quasi-ordered. This seminal result by Higman is a cornerstone of the theory of well-quasi-orderings and has found numerous applications in computer science. However, this result is based on a specific choice of ordering on words, the (scattered) subword ordering. In this paper, we describe to what extent other natural orderings (prefix, suffix, and infix) on words can be used to derive Higman-like theorems. More specifically, we are interested in characterizing \emph{languages} of words that are well-quasi-ordered under these orderings, and explore their properties and connections with other language theoretic notions. We furthermore give decision procedures when the languages are given by various computational models such as automata, context-free grammars, and automatic structures.
\end{abstract}
